Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.

### 2019-03-25 _book/electoral-modelling-draft.tex
bookdown files

  \title{Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Jeremy Forbes, Di Cook \& Rob Hyndman}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{2019-03-24}

\usepackage{booktabs}
\usepackage{amsthm}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=8pt plus 2pt minus 4pt
  \thm@postskip=\thm@preskip
}
\makeatother

@@

Estimation is done using generalized least squares (using the \texttt{spdep} \texttt{R} package \citep{spdep}).

@@

\emph{insert table here}

@@

>>>>

  \title{Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Jeremy Forbes, Di Cook \& Rob Hyndman}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{2019-03-25}

\usepackage{booktabs}
\usepackage{amsthm}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=8pt plus 2pt minus 4pt
  \thm@postskip=\thm@preskip
}
\makeatother
\usepackage{longtable}

@@

Estimation is done using feasible generalized least squares.

@@

\begin{table}[!htbp] \centering 
  \caption{Estimated model for each of the six elections.} 
  \label{} 
\tiny 
\begin{tabular}{@{\extracolsep{1pt}}lcccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{6}{c}{\textit{Dependent variable:}} \\ 
\cline{2-7} 
\\[-1.8ex] & \multicolumn{6}{c}{Two-party preferred vote in favor of the Liberal party} \\ 
 & 2001 & 2004 & 2007 & 2010 & 2013 & 2016 \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6)\\ 
\hline \\[-1.8ex] 
 $\rho$ & 0.48$^{***}$ & 0.29$^{*}$ & 0.15 & $-$0.04 & 0.15 & 0.44$^{***}$ \\ 
  & (0.16) & (0.16) & (0.17) & (0.19) & (0.17) & (0.15) \\ 
  & & & & & & \\ 
 AusCitizen & $-$1.53 & $-$0.87 & 1.00 & 0.67 & $-$2.83 & $-$2.77 \\ 
  & (2.95) & (2.90) & (2.93) & (2.57) & (2.63) & (2.51) \\ 
  & & & & & & \\ 
 Born\_Asia & $-$2.13 & $-$2.56 & $-$7.58$^{***}$ & $-$1.37 & 0.74 & 2.63 \\ 
  & (2.20) & (2.58) & (2.68) & (1.96) & (2.55) & (3.03) \\ 
  & & & & & & \\ 
 Born\_MidEast & $-$0.82 & 1.70 & 1.79 & 2.92$^{*}$ & 3.22$^{*}$ & 1.92 \\ 
  & (1.56) & (1.54) & (1.54) & (1.51) & (1.78) & (1.53) \\ 
  & & & & & & \\ 
 Born\_SE\_Europe & $-$1.63 & $-$1.25 & $-$1.72 & $-$0.20 & 0.04 & 1.02 \\ 
  & (0.99) & (1.15) & (1.34) & (1.17) & (2.05) & (2.34) \\ 
  & & & & & & \\ 
 Born\_UK & $-$1.37 & $-$0.35 & $-$0.29 & 0.29 & $-$0.87 & $-$0.49 \\ 
  & (0.98) & (1.00) & (1.05) & (0.80) & (0.90) & (1.04) \\ 
  & & & & & & \\ 
 BornElsewhere & 1.74 & 4.04 & 9.78$^{**}$ & 5.01 & $-$0.51 & $-$1.08 \\ 
  & (4.52) & (4.46) & (4.47) & (3.59) & (3.67) & (3.40) \\ 
  & & & & & & \\ 
 Buddhism & 1.51 & 1.09 & 2.41$^{*}$ & 0.90 & 2.84$^{*}$ & 1.50 \\ 
  & (1.24) & (1.25) & (1.40) & (1.25) & (1.61) & (1.94) \\ 
  & & & & & & \\ 
 Christianity & 5.17$^{**}$ & 3.08 & $-$0.60 & $-$1.93 & 3.44 & 0.53 \\ 
  & (2.52) & (2.79) & (3.44) & (3.45) & (4.19) & (4.58) \\ 
  & & & & & & \\ 
 CurrentlyStudying & $-$0.09 & 1.73 & 3.28$^{***}$ & 4.03$^{***}$ & 1.69 & $-$1.14 \\ 
  & (1.13) & (1.17) & (1.21) & (1.12) & (1.09) & (1.01) \\ 
  & & & & & & \\ 
 DeFacto & $-$11.02$^{***}$ & $-$9.73$^{***}$ & $-$8.60$^{***}$ & $-$6.55$^{***}$ & $-$5.04$^{**}$ & $-$6.42$^{***}$ \\ 
  & (2.55) & (2.62) & (2.57) & (1.90) & (2.00) & (1.69) \\ 
  & & & & & & \\ 
 DiffAddress & 6.93$^{***}$ & 5.02$^{***}$ & 4.72$^{***}$ & 3.40$^{***}$ & 4.63$^{***}$ & 3.86$^{***}$ \\ 
  & (1.50) & (1.71) & (1.60) & (0.88) & (1.03) & (0.95) \\ 
  & & & & & & \\ 
 Distributive & 1.65 & 2.76$^{**}$ & 2.88$^{**}$ & 2.25$^{**}$ & 2.75$^{**}$ & 1.67 \\ 
  & (1.17) & (1.25) & (1.30) & (1.02) & (1.13) & (1.08) \\ 
  & & & & & & \\ 
 Education & $-$3.04 & $-$0.90 & $-$1.82 & $-$0.77 & 1.94 & 0.62 \\ 
  & (2.99) & (3.04) & (3.35) & (2.89) & (2.86) & (2.23) \\ 
  & & & & & & \\ 
 Extractive & 6.81$^{***}$ & 6.09$^{***}$ & 6.97$^{***}$ & 5.81$^{***}$ & 5.31$^{***}$ & 4.69$^{***}$ \\ 
  & (1.31) & (1.42) & (1.48) & (1.16) & (1.23) & (1.36) \\ 
  & & & & & & \\ 
 FamHouseSize & $-$4.35 & $-$6.16$^{*}$ & $-$8.22$^{**}$ & $-$9.67$^{***}$ & $-$5.11$^{*}$ & $-$2.94 \\ 
  & (2.86) & (3.15) & (3.28) & (2.66) & (2.78) & (2.27) \\ 
  & & & & & & \\ 
 Incomes & 8.46$^{***}$ & 4.19 & 2.73 & 4.36$^{*}$ & 2.95 & 2.99$^{*}$ \\ 
  & (2.57) & (2.68) & (3.06) & (2.56) & (2.60) & (1.76) \\ 
  & & & & & & \\ 
 Indigenous & 4.47$^{***}$ & 6.12$^{***}$ & 5.90$^{***}$ & 6.81$^{***}$ & 5.62$^{***}$ & 3.97$^{***}$ \\ 
  & (1.38) & (1.49) & (1.49) & (1.17) & (1.35) & (1.26) \\ 
  & & & & & & \\ 
 Islam & 2.64 & 1.20 & $-$1.04 & $-$1.21 & 0.61 & 0.15 \\ 
  & (1.72) & (1.71) & (1.94) & (1.80) & (2.00) & (2.03) \\ 
  & & & & & & \\ 
 Judaism & 1.96$^{***}$ & 1.60$^{*}$ & 0.46 & 0.01 & 1.74 & 1.02 \\ 
  & (0.71) & (0.85) & (1.05) & (1.00) & (1.26) & (1.38) \\ 
  & & & & & & \\ 
 ManagerAdminClericalSales & 4.62$^{***}$ & 3.99$^{***}$ & 4.32$^{***}$ & 5.21$^{***}$ & 3.28$^{***}$ & 2.10$^{***}$ \\ 
  & (1.03) & (0.99) & (1.00) & (0.78) & (0.85) & (0.70) \\ 
  & & & & & & \\ 
 Married & $-$1.39 & 1.46 & 5.07$^{*}$ & 5.06$^{*}$ & 5.50$^{**}$ & 2.27 \\ 
  & (2.64) & (2.71) & (2.91) & (2.59) & (2.79) & (2.26) \\ 
  & & & & & & \\ 
 MedianAge & 2.94$^{*}$ & 2.28 & 2.08 & 1.86 & 3.73$^{**}$ & 1.54 \\ 
  & (1.63) & (1.88) & (2.08) & (1.65) & (1.55) & (1.30) \\ 
  & & & & & & \\ 
 NoReligion & 5.76$^{**}$ & 3.53 & 0.25 & $-$0.31 & 3.43 & 1.06 \\ 
  & (2.59) & (2.58) & (2.98) & (2.90) & (3.45) & (3.85) \\ 
  & & & & & & \\ 
 OneParent\_House & $-$0.85 & 0.11 & 0.22 & 1.66 & 1.37 & $-$1.00 \\ 
  & (1.39) & (1.46) & (1.61) & (1.38) & (1.49) & (1.34) \\ 
  & & & & & & \\ 
 OtherLanguageHome & 1.27 & $-$4.69 & $-$4.37 & $-$6.02 & $-$6.76 & $-$10.40$^{**}$ \\ 
  & (4.95) & (5.50) & (5.70) & (4.44) & (5.61) & (4.97) \\ 
  & & & & & & \\ 
 PropertyOwned & 1.76 & 0.68 & $-$1.17 & 0.78 & $-$0.92 & $-$0.76 \\ 
  & (1.38) & (1.43) & (1.49) & (1.09) & (1.21) & (1.19) \\ 
  & & & & & & \\ 
 RentLoanPrice & $-$0.70 & 1.01 & 2.12 & $-$0.54 & $-$1.54 & $-$0.31 \\ 
  & (2.02) & (1.96) & (1.96) & (1.48) & (1.68) & (1.50) \\ 
  & & & & & & \\ 
 SocialServ & 2.90$^{***}$ & 0.94 & 1.91 & 1.77 & 1.89 & 2.42$^{*}$ \\ 
  & (1.10) & (1.22) & (1.35) & (1.14) & (1.27) & (1.27) \\ 
  & & & & & & \\ 
 Transformative & 4.75$^{***}$ & 3.74$^{**}$ & 3.28$^{*}$ & 5.20$^{***}$ & 5.74$^{***}$ & 2.99$^{**}$ \\ 
  & (1.41) & (1.57) & (1.74) & (1.47) & (1.59) & (1.41) \\ 
  & & & & & & \\ 
 Unemployment & 0.36 & $-$1.22 & $-$2.42 & $-$2.88$^{**}$ & $-$5.09$^{***}$ & $-$3.00$^{**}$ \\ 
  & (1.42) & (1.54) & (1.73) & (1.41) & (1.51) & (1.28) \\ 
  & & & & & & \\ 
 Constant & 50.45$^{***}$ & 53.48$^{***}$ & 49.92$^{***}$ & 47.32$^{***}$ & 52.63$^{***}$ & 50.78$^{***}$ \\ 
  & (0.74) & (0.56) & (0.51) & (0.35) & (0.45) & (0.67) \\ 
  & & & & & & \\ 
\hline \\[-1.8ex] 
Observations & 150 & 150 & 150 & 150 & 150 & 150 \\ 
Log Likelihood & $-$399.52 & $-$405.41 & $-$418.08 & $-$398.73 & $-$403.34 & $-$398.33 \\ 
Akaike Inf. Crit. & 863.03 & 874.83 & 900.16 & 861.46 & 870.68 & 860.67 \\ 
Bayesian Inf. Crit. & 951.96 & 963.76 & 989.10 & 950.40 & 959.61 & 949.60 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{6}{l}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 & \multicolumn{6}{l}{Estimated coefficients for variable named in column one shown for} \\ 
 & \multicolumn{6}{l}{election year indicated by column heading, with estimated standard} \\ 
 & \multicolumn{6}{l}{deviation for each coefficient shown below in parenthesis. Overall} \\ 
 & \multicolumn{6}{l}{summary measures for each regression equation are provided in the} \\ 
 & \multicolumn{6}{l}{bottom panel.} \\ 
\end{tabular} 
\end{table}

@@

\begin{center}\includegraphics{electoral-modelling-draft_files/figure-latex/rhovis-1} \end{center}

The spatial autoregressive coefficient \(\rho\) is significant in only the 2001 and 2016 elections, meaning that in these years an electorate's political climate was affected by the attitudes of it's neighbours. Conversely, in the other four elections, it appears that after accounting for electoral socio-demographics, there is no spatial effect, meaning electorates effectively acted independently.

To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used. These show the direction, size and significance of an estimated effect, as the slope of the prediction line matches the estimated coefficient, and the shaded region represents a 95\% confidence band. Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners. Although extremely topical in the lead up to an election, personal, household and family incomes only affected electoral two-party preference in 2016 (see Figure \ref{fig:}).

\begin{center}\includegraphics{electoral-modelling-draft_files/figure-latex/incomes-1} \end{center}

This section summarises the main findings made about the relationships between socio-demographic chracteristics and voting behaviour.

\begin{center}\includegraphics{electoral-modelling-draft_files/figure-latex/unnamed-chunk-2-1} \end{center}
